### 2022-09-18 main/chapter_searches.tex
kern pvalue; first wave of review edits

Our task in scientific data analysis is to distil an informative
message from data and to report that message to the wider community.
From any given data there are many possible messages that could be extracted,
and those messages will have different values to different recipients of our
messages --- our data analysis is not an automatic process, but a result
of many judgements that are informed by theory, convention, and intuition.

To clarify our later description of the $\twoljets$ search in
Chapter~\ref{chapter:2ljets}, this chapter introduces the basic theory and
practice of data analysis and its manifestations in the \atlas\ SUSY community.

Theoretical data analysis in the tightly coupled fields of probability theory
and decision theory is introduced in Section~\ref{sec:searches_data_analysis}.
An attempted alternative formulation, frequentist theory, is discussed in
Section~\ref{sec:searches_frequentist} for its relevance to our methods.
Standard practice in this field of research does not strictly follow
frequentist methods, but patches up some of their pathologies with modified
approaches that we describe in Section~\ref{sec:searches_practice}.
Searches for Supersymmetric signals in \atlas\ follow a standard strategy for
their data analysis; its procedures, nomenclature, and parametric modelling
are introduced in Section~\ref{sec:searches_searches}.

\section{Data analysis}
\label{sec:searches_data_analysis}
How many jelly beans are in the jar of Figure~\ref{fig:searches_beans}?
Your answer depends not only on what you observe from the figure, but also on
what else you know about cartoon jars of beans, and on what reward is on offer.
Each answer is a decision that depends both on what the possible rewards are,
and on which numbers of beans you subjectively infer to be likely.
We shall return to this example after developing some probability and decision
theory.

@@

\TODO{cut this}
Statistical data analysis suffers from dogma --- prescriptions and recipes that
students are taught to memorize and follow without knowing their reasons or
limitations.
From a standard undergraduate physics education, my experience has been
error propagation formulae,
least-squares fitting routines,
chi-squared per degree of freedom,
and myriad statistical tests and confidence intervals.
Many of these are effective approximations, but all also fail if misused
outside their domain.%
\footnote{%
Multiplicative error propagation is an example.
For $a = \mu_a \pm \sigma_a$ and $b = \mu_b \pm \sigma_b$,
the standard perturbative approximation for their product is
\(
\widetilde{\sigma}_{ab}
= \mu_a \mu_b \sqrt{\sigma_a^2/\mu_a^2 + \sigma_b^2/\mu_b^2}
% = \sigma_a^2 \mu_b^2 + \sigma_b^2 \mu_a^2
\),
but a calculation with their moments derives the exact result
\(
\sigma_{ab}
= \sqrt{(\mu_a^2 + \sigma_a^2)(\mu_b^2 + \sigma_b^2) - \mu_a^2\mu_b^2}
% = \sigma_a^2 \mu_b^2 + \sigma_b^2 \mu_a^2 + \sigma_a^2 \sigma_b^2
\)
if $a$ and $b$ are independent.
Since $\sigma_{ab}^2 = \widetilde{\sigma}_{ab}^2 + \sigma_a^2 \sigma_b^2$, the
propagation formula works only for small relative errors.
}
All effective methods of data analysis can be derived upwards from the roots of
probability and decision theory, and as elsewhere in physics, by knowing the
derivations we can also know when each approximation should work and when it
needs replacing.

We begin with what probability is: at its most abstract, a probability a
proportion of stuff within a larger whole.
Probabilities useful because proportions have applications, and because they
can be manipulated and interrelated with a linear and intuitive algebra.
Probabilistic data analysis follows the ordinary structure of science ---
write down models that make predictions, test those predictions against data,
then update your models to refine their predictions.
For this process, likelihoods are particularly important because they describe
our predictions. But not everything is a likelihood.
The remainder of this section uses a standard example from High Energy Physics
to introduce this probabilistic structure and other parts of its technical
language.

\subsection{Probability, abstract}
A probability $\prob{b}{a}$ is a number that quantifies the proportion
of $b$ within $a$, and can be expressed as a ratio~\cite{axioms1010038}
\begin{equation}
\label{eqn:searches_prob_ratio}
\prob{b}{a} = \frac{m(b, a)}{m(a)},
\end{equation}
in which $m(x)$ is a measure (a non-negative quantity assigned to its
argument, such as mass of tea in a cup), and the comma `,' means `and'
(logical and or set intersection or any other equivalent operation).
To continue that example, $\prob{b}{a}$ could be a proportion of tea within
a spoon within a cup, where $m(b, a)$ measures the volume of tea in a
teaspoon that is partially submerged in the cup.%
\footnote{%
All probabilities are conditional in this definition; $\probc{b}$ alone does
not exist unless some conditioning information is implicit from context.%
}

Logic is an important application of probability, for which $a$ and $b$ are
interpreted as logical propositions and
$\prob{b}{a}$ is the degree to which $a$ implies $b$, with the limiting cases
that $\prob{b}{a} = 1$ when $a$ implies $b$, and $\prob{b}{a} = 0$ when $a$
contradicts $b$.
Indeed, Cox's theorem derives the laws of probability by requiring compatibility
with Boolean logic~\cite{
cox1946probability,
cox1961algebra,
garrett1998nand,
jaynes2003probability,
keynes1920treatise
}.
Probability works for logic, and it works \emph{for tea, too}.

Probability \emph{distributions} $\prob{b}{a}$ spread unit probability mass
across states $b$, and need not be random distributions of fluctuating
$b$s~\cite{jaynes2003probability,frankfurt2005on}, although random processes
are useful applications.

Two algebraic rules derive from Equation~\ref{eqn:searches_prob_ratio}:
a \textbf{product rule}
\begin{align}
\prob{c, b}{a} &= \prob{c}{b, a}\times \prob{b}{a},
\label{eqn:searches_product_rule}
\intertext{%
that describes nested proportions
(perhaps a spoon within a teacup within a sink),
and a \textbf{sum rule}%
}
\prob{c\vee b}{a} &= \prob{c}{a} + \prob{b}{a} - \prob{c, b}{a},
\label{eqn:searches_sum_rule}
\end{align}
that describes combination in a union $\vee$ that coincides with logical `or'.
The subtraction avoids double-counting of any overlap, and is often avoided by
splitting any problem into orthogonal, or disjoint states.
For disjoint cases with labels $b_i$, normalization therefore requires that
\begin{equation}
\label{eqn:searches_normalization}
\sum_i \prob{b_i}{a} = 1.
\end{equation}
These rules are basic and well known, but we state them here to support the
following statements of how probabilities can be applied for practical
results.%
\footnote{%
The formulation with sums and products is not strictly unique, but all
alternatives are equivalent up to an invertible
transformation~\cite{axioms1010038}.
This choice is usually best because it uses ordinary addition and
multiplication.
In numerical analysis, however, it is often practical to use $\log$
probabilities to avoid overflows outside the range representable by floating
point numbers, thus changing multiplication to addition and addition to
the `$\log$-add-$\exp$' operation.
}

\subsection{Probability, applied}
\label{sec:searches_probability_applied}
\begin{figure}[tp]
\centering
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/searches_abacus_likelihood.pdf}
\caption{%
Likelihood: $\mathrm{Poisson}(0\mid x)$ in
(left) $x$ and
(right) $u = \log x$%
}
\end{subfigure}
\\[.5em]
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/searches_abacus_prior.pdf}
\caption{%
Prior: exponential density in
(left) $x$ and
(right) $u = \log x$%
}
\end{subfigure}
\caption[
Likelihoods and prior densities' transformations with coordinates.
]{%
Likelihoods and prior densities' transformations with coordinates.
The $\mathrm{Poisson}(0\mid x)$ likelihood and exponential $e^{-x}\mathrm{d}x$
prior have the same form (left), but the prior density changes to preserve its unit
area relative to $u = \log x$.
On changing to $\log$ coordinates (right), the likelihood slides horizontally,
but the prior density scales to preserve area.
Both maximum likelihood points are equal, but the maximum prior density moves.
\\[.5em]
Particle lifetimes follow exponential distributions and their maximum
densities are at zero time, but it is incorrect to describe their most probable
lifetimes as zero ---
the `most probable' $\log$ lifetime is not $\log 0$, but equals the mean
lifetime.
\\[.5em]
For zero Poisson data, the maximum likelihood rate is zero, but zero might not
be the most probable rate (if, for example, an experiment has known backgrounds).
}
\label{fig:searches_abacus}
\end{figure}
% science snake
Science involves making predictions, testing those predictions against data,
and reporting what is learned from the results.
Various probabilities are involved in this prediction and learning that have
conventional names to communicate what we intend to do with them.
Most important for prediction is a \textbf{likelihood}
\begin{align}
L(a) &= \prob{b}{a} \\
\label{eqn:searches_likelihood}
\intertext{%
that explores the probabilities assigned to a fixed result $b$ from different
contexts $a$, or how well different hypotheses predicted a result.
A likelihood is clearly a probability, but it is not a probability distribution
--- only the left hand argument $b$ forms a normalized distribution.
Alternatively, fixing the context gives a \textbf{prior}%
}
\pi(b) &= \prob{b}{a}
\label{eqn:searches_prior}
\end{align}
for fixed $a$, which is a normalized probability distribution over $b$.

@@

Nonetheless, this procedure often simulates reasonable-looking test statistic
distributions, does not rely on asymptotic approximations, and reportedly
has ``good coverage properties''~\cite{
cern2011procedure,
Cranmer2006Statistical
}.
Toy simulation is an approximation, and in given circumstances may be better or
worse than the asymptotic approximations.
% Pick your poison.

>>>>

Our task in scientific data analysis is to distil useful messages from the
data, and to report that message to the wider community.
Many different messages could be extracted from given data,
and those messages will have various values to their recipients.
Data analysis is not an automatic process, but a design with many judgements
that are informed by theory, intuition, and convention to give value to their
results.

To clarify our later description of the \atlas\ SUSY $\twoljets$ search in
Chapter~\ref{chapter:2ljets}, this chapter introduces the basic theory and
practice of data analysis and its conventional form in the \atlas\ SUSY
community, parts of which can only be explained in their historical context.

Theoretical data analysis in the tightly coupled fields of probability theory
and decision theory is introduced in Section~\ref{sec:searches_data_analysis}.
An attempted alternative formulation, frequentist theory, is discussed in
Section~\ref{sec:searches_frequentist} for its relevance to our methods.
Standard practice in this field of research does not strictly follow
frequentist methods, but patches up some of their pathologies with modified
approaches that we describe in Section~\ref{sec:searches_practice}.
Searches for Supersymmetric signals in \atlas\ follow a standard strategy for
their data analysis, whose procedures, nomenclature, and parametric modelling
are introduced in Section~\ref{sec:searches_searches}.

\section{Data analysis}
\label{sec:searches_data_analysis}
How many jelly beans are in the jar of Figure~\ref{fig:searches_beans}?
Your answer depends not only on what you observe from the figure, but also on
what else you know about cartoon jars of beans, and on what reward is offered.
Each answer is a decision, which depends both on how you value the rewards,
and on which numbers of beans you subjectively infer to be probable.
We shall return to this example after developing its analysis in probability
and decision theory.

@@

Statistical data analysis suffers from dogma --- prescriptions and recipes that
one can memorize and follow while knowing neither their reasons nor their
limitations:
among these are
error propagation formulae,
least-squares fitting routines,
chi-squared per degree of freedom,
and myriad statistical tests with \pvalues\ and confidence intervals.
Each of these can be an effective, but every one fail if misused outside
its domain.%
\footnote{%
Multiplicative error propagation is an example:
for $a = \mu_a \pm \sigma_a$ and $b = \mu_b \pm \sigma_b$,
the standard rule says
\(
\widetilde{\sigma}_{ab}
= \mu_a \mu_b \sqrt{\sigma_a^2/\mu_a^2 + \sigma_b^2/\mu_b^2}
% = \sigma_a^2 \mu_b^2 + \sigma_b^2 \mu_a^2
\),
but a calculation with their moments derives the exact result
\(
\sigma_{ab}
= \sqrt{(\mu_a^2 + \sigma_a^2)(\mu_b^2 + \sigma_b^2) - \mu_a^2\mu_b^2}
% = \sigma_a^2 \mu_b^2 + \sigma_b^2 \mu_a^2 + \sigma_a^2 \sigma_b^2
\)
if $a$ and $b$ are independent;
s ince $\sigma_{ab}^2 = \widetilde{\sigma}_{ab}^2 + \sigma_a^2 \sigma_b^2$,
the propagation formula works only for small relative errors.
}
All effective methods of data analysis can be justified in probability and
decision theory.
As throughout physics, knowing a method's derivation allows one to judge when
its approximations should hold and when it leaves room for improvement.

Probabilistic data analysis follows the ordinary structure of science:
write down predictive models, test those predictions against data,
then update your models (or invent new ones) to refine their predictions
to accommodate what has been learned.
Likelihood functions are particularly important to this process because they
describe our predictions.
But as we shall see, not everything is a likelihood.

We begin with what probability is: at its most abstract, a probability a
proportion of stuff within a larger whole.
Probabilities are useful because proportions have applications
(particularly to logical reasoning with uncertainty), and because
they can be manipulated and interrelated with an intuitive linear algebra.
The remainder of this section uses a standard example from High Energy Physics
--- a histogram ---
to introduce this probabilistic structure and other parts of its technical
language.

\subsection{Probability, abstract}
A probability $\prob{b}{a}$ is a number, which quantifies the proportion
of $b$ within $a$ and can be expressed as a ratio~\cite{axioms1010038}
\begin{equation}
\label{eqn:searches_prob_ratio}
\prob{b}{a} = \frac{m(b, a)}{m(a)},
\end{equation}
in which $m(x)$ is a measure (a non-negative quantity assigned to its
argument, such as mass of tea in a cup), and the comma `,' means `and'
(logical and or set intersection or any other equivalent operation).
For example, $\prob{b}{a}$ could quantify the proportion of tea within
a spoon within a cup --- then $m(a)$ measures the content of the cup, and
$m(b, a)$ measures the amount of tea in the teaspoon that is partially
submerged in the cup.

Note that, in this definition, all probabilities are conditional.
Alone, no $\probc{b}$ exists unless some conditioning information is implicit
from context.

Logic is an important application of probability.
To apply to logical reasoning, $a$ and $b$ are logical propositions and
$\prob{b}{a}$ quantifies the degree to which $a$ implies $b$, with the
limiting cases that $\prob{b}{a} = 1$ when $a$ implies $b$,
and $\prob{b}{a} = 0$ when $a$ contradicts $b$.
Indeed, Cox's theorem shows that probability algebra is uniquely consistent
with Boolean logic~\cite{
cox1946probability,
cox1961algebra,
garrett1998nand,
jaynes2003probability,
keynes1920treatise
}.
Probability works for logic, and it works \emph{for tea, too}.

Probability \emph{distributions} $\prob{b}{a}$ spread unit probability mass
across states $b$.
They and need not be random distributions of fluctuating $b$s~\cite{
jaynes2003probability,
frankfurt2005on
},
although random processes are another useful applications.

Two algebraic rules derive from Equation~\ref{eqn:searches_prob_ratio}:
a \textbf{product rule}
\begin{align}
\prob{c, b}{a} &= \prob{c}{b, a}\times \prob{b}{a},
\label{eqn:searches_product_rule}
\intertext{%
describing nested proportions
(a spoon in a teacup in a sink),
and a \textbf{sum rule}%
}
\prob{c\;\textrm{or}\;b}{a} &= \prob{c}{a} + \prob{b}{a} - \prob{c, b}{a},
\label{eqn:searches_sum_rule}
\end{align}
describing combination in a union;
the subtraction avoids double-counting of any overlap and is usually avoided by
splitting any problem into disjoint, orthogonal states.
For disjoint cases with labels $b_i$, normalization requires that
\begin{equation}
\label{eqn:searches_normalization}
\sum_i \prob{b_i}{a} = 1.
\end{equation}
These basic rules are well known, but we state them here to support the
following statements on how probabilities can applied for practical results.%
\footnote{%
The formulation with sums and products is not strictly unique, but all
alternatives are equivalent up to an invertible
transformation~\cite{axioms1010038}.
This choice is often best because it uses ordinary addition and
multiplication, but in numerical analysis $\log$ probabilities can avoid
overflows outside the ranges representable by floating-point numbers,
thus changing multiplication to addition and addition to the
`$\log$-add-$\exp$' operation.%
}

\subsection{Probability, applied}
\label{sec:searches_probability_applied}
Science involves making predictions, testing those predictions against data,
and reporting what is learned from the results.
Various probabilities are involved in this prediction and learning that have
conventional names to communicate what we intend to do with them.
Most important for prediction is a \textbf{likelihood}
\begin{align}
L(a) &= \prob{b}{a} \\
\label{eqn:searches_likelihood}
\intertext{%
that explores the probabilities assigned to a fixed result $b$ from different
contexts $a$, or how well different hypotheses predicted a result.
A likelihood is clearly a probability, but it is not a probability distribution
--- only the left hand argument $b$ forms a normalized distribution.
Alternatively, fixing the context gives a \textbf{prior}%
}
\pi(b) &= \prob{b}{a}
\label{eqn:searches_prior}
\end{align}
for fixed $a$, which is a normalized probability distribution over $b$.

\begin{figure}[tp]
\centering
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/searches_abacus_likelihood.pdf}
\caption{%
Likelihood: $\mathrm{Poisson}(0\mid x)$ in
(left) $x$ and
(right) $u = \log x$%
}
\end{subfigure}
\\[.5em]
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/searches_abacus_prior.pdf}
\caption{%
Prior: exponential density in
(left) $x$ and
(right) $u = \log x$%
}
\end{subfigure}
\caption[
Likelihoods and prior densities' transformations with coordinates.
]{%
Likelihoods and prior densities' transformations with coordinates.
The $\mathrm{Poisson}(0\mid x)$ likelihood and exponential $e^{-x}\mathrm{d}x$
prior have the same form (left), but the prior density changes to preserve its unit
area relative to $u = \log x$.
On changing to $\log$ coordinates (right), the likelihood slides horizontally,
but the prior density scales to preserve area.
Both maximum likelihood points are equivalent, but the maximum prior density
moves.
\\[.5em]
Particle lifetimes follow exponential distributions and their maximum
densities are at zero time, but it is incorrect to describe their most probable
lifetimes as zero ---
the `most probable' $\log$ lifetime is not $\log 0$, but turns out to coincide
with the mean lifetime.
\\[.5em]
For zero Poisson data, the maximum likelihood rate is zero, but zero might not
be the most probable rate (if, for example, an experiment has known backgrounds).
}
\label{fig:searches_abacus}
\end{figure}

@@

Nonetheless, this procedure often simulates reasonable-looking test statistic
distributions, does not rely on asymptotic approximations, and purportedly
has ``good coverage properties''~\cite{
cern2011procedure,
Cranmer2006Statistical
}.
Toy simulation is an approximation, and in given circumstances may be better or
worse than the asymptotic approximations.
% Pick your poison.
